---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r cache=TRUE}
# file name
file_name <- "activity.zip"
# read file and save the data in a data frame, named as 'df'
df <- read.csv(unz(file_name, "activity.csv"), header = TRUE, sep = ",")
# display the head data in the data frame
head(df)
```


## What is mean total number of steps taken per day?
```{r cache=TRUE}
knitr::opts_chunk$set(fig.path = "figure/")
# find data without missing
isGoodData <- !is.na(df[, 1]) 
# group the data according to date
dayNumSteps <- split(df[isGoodData, 1], df[isGoodData, 2])
# calculate sum of the number of steps taken each day
totNumSteps <- sapply(dayNumSteps, sum)
# make a histogram of the total number steps taken each day
hist(totNumSteps, main = "Histogram of the total number steps")
# calculate the mean and median
goodTotNumSteps <- numeric() 
iCount <- 0
for (i in 1:length(totNumSteps)) {
    if (length(dayNumSteps[[i]]) != 0) {
        iCount <- iCount + 1
        goodTotNumSteps[iCount] <- totNumSteps[[i]]
    }
}
# display total number of steps in days the date are taken
goodTotNumSteps
# mean
mean(goodTotNumSteps)
# median
median(goodTotNumSteps)
```


## What is the average daily activity pattern?
```{r cache=TRUE}
# group the data according to interval
intervalNumSteps <- split(df[isGoodData, 1], df[isGoodData, 3])
# calculate sum of the number of steps taken each day
totIntervalNumSteps <- sapply(intervalNumSteps, sum)
averageIntervalNumSteps <- numeric(length = length(totIntervalNumSteps))
for (i in 1:length(totIntervalNumSteps)) {
    averageIntervalNumSteps[i] <- totIntervalNumSteps[i] / length(intervalNumSteps[[i]])
}
# avarage number of steps taken
averageIntervalNumSteps
# Calculate time axis
timeVal <- numeric(length = 288)
for (i in 1:288) {
    timeFormat <- strptime(sprintf("%04.0f", df[i, 3]), format = "%H%M")
    timeVal[i] <- timeFormat$hour * 60 + timeFormat$min
}
# time axis
timeVal
```

```{r cache=TRUE}
knitr::opts_chunk$set(fig.path = "figure/")
# plot the data
plot(timeVal, averageIntervalNumSteps,
     main = "Averaged across all days",
     xlab = "", ylab = "Average number of steps",
     col = "black",
     type = "l",
     xaxt = "n")
xsticks <- c(0 * 60, 5 * 60, 10 * 60, 15 * 60, 20 * 60)
days <- c("00:00", "05:00", "10:00", "15:00", "20:00") 
axis(side = 1, at = xsticks, labels = days, tck = -0.05)
# interval of maximum number of steps
intervalMax <- timeVal[averageIntervalNumSteps == max(averageIntervalNumSteps)]
sprintf("%02.0f:%02.0f", intervalMax %/% 60, intervalMax %% 60)
```


## Imputing missing values
```{r cache=FALSE}
# number of missing data
sum(is.na(df[, 1])) 
# impute the missing data by mean (the mean for 5-minute interval)
new_df <- data.frame(df)
for (i in 1:nrow(new_df)) {
    if (is.na(new_df[i, 1]) == TRUE) {
        new_df[i, 1] <- as.integer(round(averageIntervalNumSteps[((i %% 288) == 0) * 288 + (i %% 288)]))
    }
}
# display head of imputed data
head(new_df)
```

```{r cache=FALSE}
knitr::opts_chunk$set(fig.path = "figure/")
# group the data according to date
new_dayNumSteps <- split(new_df[, 1], new_df[, 2])
# calculate sum of the number of steps taken each day
new_totNumSteps <- sapply(new_dayNumSteps, sum)
# make a histogram of the total number steps taken each day
hist(new_totNumSteps, main = "Histogram of the total number steps (imputed)")
# mean
mean(new_totNumSteps)
# median
median(new_totNumSteps)
```

Compare with the mean and median in the first part, they slightly different (smaller than the values in the first part).

```{r cache=TRUE}
knitr::opts_chunk$set(fig.path = "figure/")
# group the data according to interval
new_intervalNumSteps <- split(new_df[, 1], new_df[, 3])
# calculate average of the number of steps taken each day
new_averageIntervalNumSteps <- sapply(new_intervalNumSteps, mean)
# plot the data
matplot(timeVal, 
       cbind(averageIntervalNumSteps, new_averageIntervalNumSteps),
       main = "Averaged across all days (imputed)",
       xlab = "", ylab = "Average number of steps (imputed)",
       col = c("black", "red"),
       lwd = c(4, 1),
       lty = c(1, 1),
       type = c("l", "l"),
       xaxt = "n")
xsticks <- c(0 * 60, 5 * 60, 10 * 60, 15 * 60, 20 * 60)
days <- c("00:00", "05:00", "10:00", "15:00", "20:00") 
axis(side = 1, at = xsticks, labels = days, tck = -0.05)
legend("topright", legend = c("Missing", "Imputed"), col = c("black", "red"), lty = 1)

# interval of maximum number of steps
new_intervalMax <- timeVal[new_averageIntervalNumSteps == max(new_averageIntervalNumSteps)]
sprintf("%02.0f:%02.0f", intervalMax %/% 60, intervalMax %% 60)
```
Since imputation of data uses the mean of 5-minute interval, the plot of averaged number across all days is insignificantly changed.

## Are there differences in activity patterns between weekdays and weekends?
```{r cache=TRUE}
# add a column of day
new_df$day <- weekdays(strptime(as.character(new_df[, 2]), format = "%Y-%m-%d"))
for (i in 1:nrow(new_df)) {
    if ((new_df$day[i] != "Saturday") & (new_df$day[i] != "Sunday"))
        new_df$dayType[i] <- "weekday"
    else
        new_df$dayType[i] <- "weekend"
}
# factor variable
new_df$dayType <- factor(new_df$dayType)
```

```{r cache=TRUE}
# split data
weekdayData <- split(new_df[new_df[, 5] == "weekday", 1], new_df[new_df[, 5] == "weekday", 3])
weekendData <- split(new_df[new_df[, 5] == "weekend", 1], new_df[new_df[, 5] == "weekend", 3])
# calculate average of the number of steps taken each day
wkd_averageIntervalNumSteps <- sapply(weekdayData, mean)
wke_averageIntervalNumSteps <- sapply(weekendData, mean)
```

```{r cache=FALSE}
knitr::opts_chunk$set(fig.path = "figure/")
par(mfrow = c(2, 1), 
    oma = c(0, 0, 0, 0),
    omd = c(0, 0, 0, 0),
    omi = c(0, 0, 0, 0),
    mai = c(0.8, 0.8, 0.2, 0.2))

# Plot 2.1: Plot of weekend data
plot(timeVal, wke_averageIntervalNumSteps,
     main = "Averaged across all weekends",
     xlab = "", ylab = "Average number of steps",
     col = "black",
     type = "l",
     xaxt = "n")
xsticks <- c(0 * 60, 5 * 60, 10 * 60, 15 * 60, 20 * 60)
days <- c("00:00", "05:00", "10:00", "15:00", "20:00") 
axis(side = 1, at = xsticks, labels = days, tck = -0.05)

# Plot 4.2: Plot of weekday data
plot(timeVal, wkd_averageIntervalNumSteps,
     main = "Averaged across all weekdays",
     xlab = "", ylab = "Average number of steps",
     col = "black",
     type = "l",
     xaxt = "n")
xsticks <- c(0 * 60, 5 * 60, 10 * 60, 15 * 60, 20 * 60)
days <- c("00:00", "05:00", "10:00", "15:00", "20:00") 
axis(side = 1, at = xsticks, labels = days, tck = -0.05)
```